cardinalR: Generating interesting high-dimensional data structures

### 2025-06-24 paper-cardinalR.tex
Updated \frac

Branch \(1\) is generated from \(X_1 \sim U(0, 1)\), and \(X_2 = 0.1X_1 + X_1^2 + \epsilon\), where \(\epsilon \sim U(0, 0.05)\). This produces a gently upward-curving parabola in the right half-plane. Branch \(2\) is generated from \(X_{1} \sim U(-1, 0)\), and \(X_2 = 0.1X_{1} - 2X_{1}^2 + \epsilon\), where \(\epsilon \sim U(0, 0.05)\). This creates a steeper, leftward-facing curve in the left half-plane. Branches \(3\) to \(k\) are added iteratively. Each new branch \(i\) begins at a randomly chosen point within a restricted horizontal band: \(X_{1} \in [-0.15, 0.15]\) (to ensure connectivity with earlier branches), spans a unit-length interval on \(X_{1}\):\(X_1 \in [x_{\text{start}}, x_{\text{start}} + 1]\), and has the structure:

\[
X_2 = 0.1 \cdot X_1 - s_i \cdot (X_1^2 - x_{\text{start}}) + y_{\text{start}}, \quad s_i \in \{-2, -1.5, -1, -0.5, 0, 0.5, 1.5\}.
\]

Here, \(s_i\) is a sampled scale factor (slope-like term) controlling curvature, and \((x_{\text{start}}, y_{\text{start}})\) is the starting point from the existing structure.

This construction yields \(k\) spatially connected, nonlinear branches with varying curvature.

For \(p > 2\), Gaussian noise \(X_j \sim N(0, 0.05^2)\) is added to embed the \(2\text{-}D\) branches into \(p\text{-}D\), where \(j = 3, \dots, p\).

@@

For branch \(i\), \(n_i\) points are generated, where \(\sum_{i=1}^k n_i = n\). The structure follows: \(X_{i1} \sim U(0, 2)\), and \(X_{i2} = -s_i \cdot x_{i1} + \epsilon, \quad \epsilon \sim U(0, 0.5)\), where \(s_i\) is a scale factor controlling the slope. When \(k \leq \binom{p}{2}\), \(s_i = 1\). When sampling with replacement, \(s_i\) is drawn from the set \({1, 1.5, 2, \dots, 8}\) in increments of \(0.5\). For \(p > 2\), the remaining dimensions contain independent Gaussian noise: \(X_j \sim N(0, 0.1^2)\), for \(j \notin \{i1, i2\}\).

@@

Each branch contains \(n_i\) points such that \(\sum_{i=1}^k n_i = n\), where the vector \((n_1, \dots, n_k)\) is randomly drawn using the helper function \texttt{gen\_nsum()} to partition \(n\). Within its assigned subspace \((X_{i1}, X_{i2})\), branch \(i\) is defined by: \(X_{i1} \sim U(0, 2)\), and \(X_{i2} = -s_i \cdot X_{i1}^2 + \epsilon, \quad \epsilon \sim U(0, 0.5)\). This forms a smooth downward-opening parabola in the plane defined by \((X_{i1}, X_{i2})\), with the degree of curvature controlled by \(s_i\). For \(p > 2\), all remaining dimensions are independent Gaussian noise: \(X_j \sim N(0, 0.1^2), \quad \text{for } j \notin \{i1, i2\}\).

@@

The Mobius strip is generated using \texttt{geozoo::mobius(n,\ p\ =\ 3)\$points}, which samples \(n\) points from a parametrization of the surface defined by angle \(\theta\) and strip width \(w\). The cartesian coordinates are \(X_1 = \left(1 + \frac{w}{2} \cos(\theta/2)\right) \cos(\theta)\), \(X_2 = \left(1 + \frac{w}{2} \cos(\theta/2)\right) \sin(\theta)\), and \(X_3 = \frac{w}{2} \sin(\theta/2)\). This maps a \(2\text{-}D\) band with a half-twist into \(3\text{-}D\) space, forming a non-orientable one-sided surface. For \(p > 3\), Gaussian noise \(X_j \sim N(0, 0.05^2)\) is added to embed the \(3\text{-}D\) Mobius into \(p\text{-}D\), where \(j = 4, \dots, p\).

@@

Let \(X_1, X_2, \dots, X_p\) denote the coordinates of the generated points. The final dimension \(X_p\) encodes the height of each point and is drawn from an exponential distribution capped at \(h\): \(X_p = z \sim \min\left(\text{Exp}(\lambda = 2/h),\ h\right)\). Let \(r_x(z)\) and \(r_y(z)\) denote the half-widths of the rectangular cross-section at height \(z\): \(r_x(z) = r_t + (l_x - r_t) \cdot \frac{z}{h}\), \(r_y(z) = r_t + (l_y - r_t) \cdot \frac{z}{h}\). The first three coordinates are then defined as: \(X_1 \sim U(-r_x(z),\ r_x(z))\), \(X_2 \sim U(-r_y(z),\ r_y(z))\), and \(X_3 \sim U(-r_x(z),\ r_x(z))\). For \(p > 3\), the remaining \(p - 4\) dimensions (i.e., \(X_4\) to \(X_{p-1}\)) are tapered toward zero with decreasing height and generated as: \(X_j \sim U(-0.1, 0.1) \cdot \left(1 - \frac{z}{h}\right), \quad \text{for } j = 4, \dots, p-1\). This tapering ensures that the structure narrows in all directions as it approaches the tip.

@@

Let \(X_1, X_2, \dots, X_p\) denote the coordinates of the generated points. The final coordinate \(X_p\) encodes the height \(z\) of each point and is drawn from an exponential distribution capped at \(h\): \(X_p = z \sim \min\left(\text{Exp}(\lambda = 2/h),\ h\right)\). Let \(r(z)\) denote the scaling factor (distance from the origin to triangle vertices) at height \(z\): \(r(z) = r_t + (l - r_t) \cdot \frac{z}{h}\).

A point in the triangle at height \(z\) is generated using barycentric coordinates \((u, v)\) to ensure uniform sampling within the triangular cross-section: \(u, v \sim U(0, 1), \quad \text{if } u + v > 1: u \leftarrow 1 - u,\ v \leftarrow 1 - v\). The first three coordinates (triangle plane) are then: \(X_1 = r(z)(1 - u - v)\), \(X_2 = r(z) \cdot u\), and \(X_3 = r(z) \cdot v\). For dimensions \(j = 4, \dots, p-1\), values are tapered linearly toward zero near the tip: \(X_j \sim U(-0.1, 0.1) \cdot \left(1 - \frac{z}{h}\right)\).

@@

The radius at height \(z\) scales linearly from zero (tip) to the base radius \(r_b\): \(r(z) = r_b \cdot \left(1 - \frac{z}{h}\right)\).

Each point is placed within a regular hexagon in the plane \((X_1, X_2)\), using a randomly chosen hexagon sector angle \(\theta \in \{0, \frac{\pi}{3}, \frac{2\pi}{3}, \pi, \frac{4\pi}{3}, \frac{5\pi}{3}\}\) and a uniformly random radial scaling factor: \(\theta \sim \text{Uniform sample from 6 hexagon angles}\),
\(r_{\text{point}} \sim \sqrt{U(0, 1)}\). Then, the first two coordinates are: \(X_1 = r(z) \cdot r_{\text{point}} \cdot \cos(\theta)\), and \(X_2 = r(z) \cdot r_{\text{point}} \cdot \sin(\theta)\). For dimensions \(j = 3, \dots, p-1\), values are tapered toward zero near the tip: \(X_j \sim U(-0.1, 0.1) \cdot \left(1 - \frac{z}{h} \right)\).

@@

Let \(C_1, C_2, \dots, C_{p+1}\) denote the corner vertices of a \(p\text{-}D\) simplex. At each iteration \(i = 1, \dots, n\), a new point is computed by taking the midpoint between the previous point \(T_{i-1}\) and a randomly selected vertex \(C_k\): \(T_i = \frac{1}{2}(T_{i-1} + C_k), \quad C_k \in \{C_1, \dots, C_{p+1}\}\). This recursive midpoint rule generates self-similar patterns with systematic voids (holes) between clusters of points. The points remain bounded inside the convex hull of the simplex. The final output is a \(n \times p\) matrix where each row represents a point: \(X = \{T_1, T_2, \dots, T_n\}, \quad X \in \mathbb{R}^{n \times p}\).

@@

The \(3\text{-}D\) geometry is constructed by introducing a latent parameter, \(\theta \sim U\left(-\frac{3\pi}{2}, \frac{3\pi}{2}\right)\). This parameter controls the curvature of the manifold. The first three dimensions form the S-curve structure: \(X_1 = \sin(\theta)\), \(X_2 \sim U(0, 2)\), \(X_3 = \text{sign}(\theta) \cdot (\cos(\theta) - 1)\). This configuration creates a horizontally curled shape in \((X_1, X_3)\), with additional band thickness in the \(X_2\) direction. For \(p > 3\), additional noise dimensions are appended introducing structured, wavy noise.

@@

A latent angle variable \(\theta\) is uniformly sampled from the interval \([0, 2\pi]\). Coordinates in the first two dimensions represent a perfect circle on the plane: \(X_1 = \cos(\theta), \quad X_2 = \sin(\theta)\). For dimensions \(X_3\) through \(X_p\), sinusoidal transformations of the angle \(\theta\) are introduced. The first component is a scaling factor that decreases with the dimension index, defined as \(\text{scale}_j = \sqrt{(0.5)^{j-2}}\) for \(j = 3, \dots, p\). The second component is a phase shift that is proportional to the dimension index, specifically designed to decorrelate the curves, given by the formula \(\phi_j = (j - 2) \cdot \frac{\pi}{2p}\). Each additional dimension is computed as: \(X_j = \text{scale}_{j} \cdot \sin(\theta + \phi_j), \quad j = 3, \dots, p\).

@@

A latent angle variable \(\theta\) is uniformly sampled from the interval \([0, 2\pi]\). The first three dimensions define a non-circular closed curve, referred to as a ``curvy cycle.'' In this configuration, \(X_1 = \cos(\theta)\) represents horizontal oscillation, while \(X_2 = \sqrt{3}/3 + \sin(\theta)\) introduces a vertical offset to avoid centering the curve at the origin. Additionally, \(X_3 = \frac{1}{3} \cdot \cos(3\theta)\) introduces a third harmonic perturbation that intricately folds the curve three times along its path, creating a unique and complex shape that oscillates in both dimensions while incorporating the effects of the harmonic perturbation.

Together, these define a periodic, non-trivial, closed curve in \(3\text{-}D\) with internal folds that produce a more complex geometry than a standard circle or ellipse. For dimensions \(X_4\) through \(X_p\), additional structured variability is introduced through decreasing amplitude scaling and phase-shifted sine waves. The scaling factor is defined as \(\text{scale}_j = \sqrt{(0.5)^{j-3}}\) for \(j\) ranging from \(4\) to \(p\), which means that the amplitude decreases as the dimension increases. Each dimension \(X_j\) is then calculated using the formula \(X_j = \text{scale}_j \cdot \sin(\theta + \phi_j)\), where the phase shift \(\phi_j\) is given by \(\phi_j = (j - 2) \cdot \frac{\pi}{2p}\).

@@

Each observation lies on the surface of a sphere in \(3\text{-}D\), constructed by generating \(u \sim U(-1, 1)\) which represents the cosine of the polar angle \(\phi\) and \(\theta \sim U(0, 2\pi)\) which represents the azimuthal angle. The corresponding Cartesian coordinates are calculated as; \(X_1 = r \cdot \sqrt{1 - u^2} \cdot \cos(\theta)\), \(X_2 = r \cdot \sqrt{1 - u^2} \cdot \sin(\theta)\), \(X_3 = r \cdot u\), which gives points uniformly distributed on the surface of a \(3\text{-}D\) sphere (not within). For \(p > 3\), additional dimensions \(X_4\) to \(X_p\) are generated as low-variance Gaussian noise: \(X_j \sim N(0, 0.05^2),\text{ for }j = 4, \dots, p\).

@@

Each point on the sphere is computed using the spherical-to-Cartesian transformation: \(X_1 = \sin(\phi) \cdot \cos(\theta)\), \(X_2 = \sin(\phi) \cdot \sin(\theta)\), \(X_3 = \cos(\phi)\). For \(p > 3\), the manifold is embedded in a higher-dimensional space by adding Gaussian noise dimensions: \(X_j \sim N(0, 0.05^2),\text{ for }j = 4, \dots, p\).

@@

Each observation is situated on a restricted \(4\text{-}D\) spherical surface, defined by spherical coordinates. The azimuthal angle \(\theta_1 \sim U(0, \pi)\) in the \((x_1, x_2)\) plane, while the elevation angle \(\theta_2 \sim U(0, \pi)\) in the \((x_2, x_3)\) plane. Additionally, \(\theta_3 \sim U(0, \pi/2)\) in the \((x_3, x_4)\) plane, ensuring that the points remain restricted to a hemisphere. The coordinates are transformed into \(4\text{-}D\) Cartesian space: \(X_1 = \sin(\theta_1) \cdot \cos(\theta_2)\), \(X_2 = \sin(\theta_1) \cdot \sin(\theta_2)\), \(X_3 = \cos(\theta_1) \cdot \cos(\theta_3)\), \(X_4 = \cos(\theta_1) \cdot \sin(\theta_3)\). This produces points on one side of a \(4\text{-}D\) unit sphere, effectively generating a \(4\text{-}D\) hemisphere. For \(p > 4\), additional dimensions \(X_5\) to \(X_p\) are added as low-variance Gaussian noise: \(X_j \sim N(0, 0.05^2),\text{ for }j = 5, \dots, p\).

@@

The first two dimensions define a crescent-shaped curve using a semi-circular arc. Let \(\theta \in [\pi/6, 2\pi]\) be a sequence of \(n\) evenly spaced angles. The corresponding \(2\text{-}D\) coordinates are defined by: \(X_1 = \cos(\theta)\), \(X_2 = \sin(\theta)\). For \(p > 2\), additional dimensions \(X_3\) to \(X_p\) are generated using \texttt{gen\_wavydims1()}: \(X_j = s_j \cdot \theta + \varepsilon_j,\text{ where }\varepsilon_j \sim N(0, 0.5^2),\text{ and }s_j\text{ is a random scale}\). These noise dimensions are nonlinearly correlated with \(\theta\), producing structured, wave-like variation that aligns with the progression along the crescent curve.

@@

Cartesian coordinates from spherical conversion: \(X_1 = \sin(\phi) \cdot \cos(\theta)\), \(X_2 = \sin(\phi) \cdot \sin(\theta)\), \(X_3 = \cos(\phi) + \varepsilon\), where \(\varepsilon \sim U(-0.5, 0.5)\) introduces vertical jitter, and \(X_4 = \theta / \max(\theta)\): a normalized progression along the spiral path. This generates a spherical spiral curve embedded in \(4\text{-}D\) space, combining both circular and vertical movement, with gentle curvature and non-linear progression.

If \(p > 4\), the function appends structured, non-linear noise via \texttt{gen\_wavydims2()}. These dimensions are functions of the first coordinate \(X_1\), introducing dependencies that preserve some geometric coherence. Each added dimension follows the form: \(X_j = s_j \cdot (-1)^{\lfloor j/2 \rfloor} \cdot X_1^{a_j} + \eta_j\), where \(a_j \in \{2, 3, 4, 5\}\) is a randomly chosen polynomial power, \(s_j \sim U(0.5, 2)\) is a scale factor, \(\eta_j \sim U(-\sigma_j, 2\sigma_j)\), with \(\sigma_j \sim U(0, 0.05)\), adds mild randomness.

@@

The first four coordinates follow a \(3\text{-}D\) helix with an additional oscillating component. Let \(\theta \in [0, \frac{5\pi}{4}]\) be a sequence of angles controlling rotation around a circle. Cartesian coordinates; \(X_1 = \cos(\theta)\): circular trajectory along the x-axis, \(X_2 = \sin(\theta)\): circular trajectory along the y-axis, \(X_3 = 0.05 \cdot \theta + \varepsilon_3\), with \(\varepsilon_3 \sim U(-0.5, 0.5)\): linear progression (height) with vertical jitter, simulating a helix, and \(X_4 = 0.1 \cdot \sin(\theta)\): oscillates with \(\theta\), representing a periodic ``wobble'' along the fourth dimension. This results in a helical spiral winding upward along the third axis, with gentle sinusoidal fluctuation in the fourth dimension. When \(p > 4\), the remaining \(p - 4\) dimensions are populated using \texttt{gen\_noisedims()}. These are independent Gaussian noise dimensions: \(X_j \sim N(0, 0.05^2), \quad j = 5, \dots, p\).

@@

The shape is defined by parameter \(\theta \in [0, 2\pi \cdot \text{spins}]\), controlling the angular progression of the spiral. The Archimedean spiral in the horizontal plane is represented by; \(X_1 = \theta \cdot \cos(\theta)\) for radial expansion in x, and \(X_2 = \theta \cdot \sin(\theta)\) for radial expansion in y. The growth pattern resembles a cone, with the height increasing according to \(X_3 = 2 \cdot \theta / \max(\theta) + \varepsilon_3\), with \(\varepsilon_3 \sim U(-0.1, 0.6).\) Spiral modulation in the fourth dimension is represented by \(X_4 = \theta \cdot \sin(2\theta) + \varepsilon_4\), with \(\varepsilon_4 \sim U(-0.1, 0.6)\) which simulates a twisting helical component in a non-radial dimension. This results in a \(3\text{-}D\) spiral surface expanding upward and outward like a cone, with an oscillatory fourth dimension capturing spiral irregularities. For \(p > 4\), the dataset includes isotropic Gaussian noise in the remaining dimensions via \texttt{gen\_noisedims()}: \(X_j \sim N(0, 0.05^2), \quad j = 5, \dots, p\).

@@

The \(X_{1} \sim U(0.1, 2)\): base variable (avoids zero to prevent division errors), \(X_{3} \sim U(0.1, 0.8)\): independent auxiliary variable, \(X_{2} = \frac{hc}{X_{1}} + \text{nonfac} \cdot \sin(X_{1})\): non-linear combination of hyperbolic and sinusoidal transformations, creating sharp curvature and oscillation, and \(X_{4} = \cos(\pi \cdot X_{1}) + \varepsilon\), with \(\varepsilon \sim U(-0.1, 0.1)\): additional nonlinear variation based on cosine, simulating more subtle periodic structure. These transformations together result in a non-linear surface warped in multiple ways: sharp vertical shifts due to inverse terms, smooth waves from sine and cosine, and additional jitter. If \(p > 4\), the remaining dimensions are populated with Gaussian noise using \texttt{gen\_noisedims()}: \(X_j \sim N(0, 0.05^2), \quad j = 5, \dots, p\).

>>>>

Branch \(1\) is generated from \(X_1 \sim U(0, 1)\), and \(X_2 = 0.1X_1 + X_1^2 + \epsilon\), where \(\epsilon \sim U(0, 0.05)\). This produces a gently upward-curving parabola in the right half-plane. Branch \(2\) is generated from \(X_{1} \sim U(-1, 0)\), and \(X_2 = 0.1X_{1} - 2X_{1}^2 + \epsilon\), where \(\epsilon \sim U(0, 0.05)\). This creates a steeper, leftward-facing curve in the left half-plane. Branches \(3\) to \(k\) are added iteratively. Each new branch \(i\) begins at a randomly chosen point within a restricted horizontal band: \(X_{1} \in [-0.15, 0.15]\) (to ensure connectivity with earlier branches), spans a unit-length interval on \(X_{1}\):\(X_1 \in [x_{\text{start}}, x_{\text{start}} + 1]\), and has the structure: \(X_2 = 0.1X_1 - s_i(X_1^2 - x_{\text{start}}) + y_{\text{start}}, s_i \in \{-2, -1.5, -1, -0.5, 0, 0.5, 1.5\}\). Here, \(s_i\) is a sampled scale factor (slope-like term) controlling curvature, and \((x_{\text{start}}, y_{\text{start}})\) is the starting point from the existing structure. This construction yields \(k\) spatially connected, nonlinear branches with varying curvature. For \(p > 2\), Gaussian noise \(X_j \sim N(0, 0.05^2)\) is added to embed the \(2\text{-}D\) branches into \(p\text{-}D\), where \(j = 3, \dots, p\).

@@

For branch \(i\), \(n_i\) points are generated, where \(\sum_{i=1}^k n_i = n\). The structure follows: \(X_{i1} \sim U(0, 2)\), and \(X_{i2} = -s_iX_{i1} + \epsilon, \quad \epsilon \sim U(0, 0.5)\), where \(s_i\) is a scale factor controlling the slope. When \(k \leq \binom{p}{2}\), \(s_i = 1\). When sampling with replacement, \(s_i\) is drawn from the set \({1, 1.5, 2, \dots, 8}\) in increments of \(0.5\). For \(p > 2\), the remaining dimensions contain independent Gaussian noise: \(X_j \sim N(0, 0.1^2)\), for \(j \notin \{i1, i2\}\).

@@

Each branch contains \(n_i\) points such that \(\sum_{i=1}^k n_i = n\), where the vector \((n_1, \dots, n_k)\) is randomly drawn using the helper function \texttt{gen\_nsum()} to partition \(n\). Within its assigned subspace \((X_{i1}, X_{i2})\), branch \(i\) is defined by: \(X_{i1} \sim U(0, 2)\), and \(X_{i2} = -s_iX_{i1}^2 + \epsilon, \quad \epsilon \sim U(0, 0.5)\). This forms a smooth downward-opening parabola in the plane defined by \((X_{i1}, X_{i2})\), with the degree of curvature controlled by \(s_i\). For \(p > 2\), all remaining dimensions are independent Gaussian noise: \(X_j \sim N(0, 0.1^2), \quad \text{for } j \notin \{i1, i2\}\).

@@

The Mobius strip is generated using \texttt{geozoo::mobius(n,\ p\ =\ 3)\$points}, which samples \(n\) points from a parametrization of the surface defined by angle \(\theta\) and strip width \(w\). The cartesian coordinates are \(X_1 = \left(1 + w/2 \cos(\theta/2)\right) \cos(\theta)\), \(X_2 = \left(1 + w/2 \cos(\theta/2)\right) \sin(\theta)\), and \(X_3 = w/2 \sin(\theta/2)\). This maps a \(2\text{-}D\) band with a half-twist into \(3\text{-}D\) space, forming a non-orientable one-sided surface. For \(p > 3\), Gaussian noise \(X_j \sim N(0, 0.05^2)\) is added to embed the \(3\text{-}D\) Mobius into \(p\text{-}D\), where \(j = 4, \dots, p\).

@@

Let \(X_1, X_2, \dots, X_p\) denote the coordinates of the generated points. The final dimension \(X_p\) encodes the height of each point and is drawn from an exponential distribution capped at \(h\): \(X_p = z \sim \min\left(\text{Exp}(\lambda = 2/h),\ h\right)\). Let \(r_x(z)\) and \(r_y(z)\) denote the half-widths of the rectangular cross-section at height \(z\): \(r_x(z) = r_t + (l_x - r_t)z/h\), \(r_y(z) = r_t + (l_y - r_t)z/h\). The first three coordinates are then defined as: \(X_1 \sim U(-r_x(z),\ r_x(z))\), \(X_2 \sim U(-r_y(z),\ r_y(z))\), and \(X_3 \sim U(-r_x(z),\ r_x(z))\). For \(p > 3\), the remaining \(p - 4\) dimensions (i.e., \(X_4\) to \(X_{p-1}\)) are tapered toward zero with decreasing height and generated as: \(X_j \sim U(-0.1, 0.1)\left(1 - z/h\right), \quad \text{for } j = 4, \dots, p-1\). This tapering ensures that the structure narrows in all directions as it approaches the tip.

@@

Let \(X_1, X_2, \dots, X_p\) denote the coordinates of the generated points. The final coordinate \(X_p\) encodes the height \(z\) of each point and is drawn from an exponential distribution capped at \(h\): \(X_p = z \sim \min\left(\text{Exp}(\lambda = 2/h),\ h\right)\). Let \(r(z)\) denote the scaling factor (distance from the origin to triangle vertices) at height \(z\): \(r(z) = r_t + (l - r_t)z/h\).

A point in the triangle at height \(z\) is generated using barycentric coordinates \((u, v)\) to ensure uniform sampling within the triangular cross-section: \(u, v \sim U(0, 1), \quad \text{if } u + v > 1: u \leftarrow 1 - u,\ v \leftarrow 1 - v\). The first three coordinates (triangle plane) are then: \(X_1 = r(z)(1 - u - v)\), \(X_2 = r(z)u\), and \(X_3 = r(z)v\). For dimensions \(j = 4, \dots, p-1\), values are tapered linearly toward zero near the tip: \(X_j \sim U(-0.1, 0.1)\left(1 - z/h\right)\).

@@

The radius at height \(z\) scales linearly from zero (tip) to the base radius \(r_b\): \(r(z) = r_b\left(1 - z/h\right)\).

Each point is placed within a regular hexagon in the plane \((X_1, X_2)\), using a randomly chosen hexagon sector angle \(\theta \in \{0, \pi/3, 2\pi/3, \pi, 4\pi/3, 5\pi/3\}\) and a uniformly random radial scaling factor: \(\theta \sim \text{Uniform sample from 6 hexagon angles}\),
\(r_{\text{point}} \sim \sqrt{U(0, 1)}\). Then, the first two coordinates are: \(X_1 = r(z)r_{\text{point}}\cos(\theta)\), and \(X_2 = r(z)r_{\text{point}}\sin(\theta)\). For dimensions \(j = 3, \dots, p-1\), values are tapered toward zero near the tip: \(X_j \sim U(-0.1, 0.1)\left(1 - z/h \right)\).

@@

Let \(C_1, C_2, \dots, C_{p+1}\) denote the corner vertices of a \(p\text{-}D\) simplex. At each iteration \(i = 1, \dots, n\), a new point is computed by taking the midpoint between the previous point \(T_{i-1}\) and a randomly selected vertex \(C_k\): \(T_i = 1/2(T_{i-1} + C_k), \quad C_k \in \{C_1, \dots, C_{p+1}\}\). This recursive midpoint rule generates self-similar patterns with systematic voids (holes) between clusters of points. The points remain bounded inside the convex hull of the simplex. The final output is a \(n \times p\) matrix where each row represents a point: \(X = \{T_1, T_2, \dots, T_n\}, \quad X \in \mathbb{R}^{n \times p}\).

@@

The \(3\text{-}D\) geometry is constructed by introducing a latent parameter, \(\theta \sim U\left(-3\pi/2, 3\pi/2\right)\). This parameter controls the curvature of the manifold. The first three dimensions form the S-curve structure: \(X_1 = \sin(\theta)\), \(X_2 \sim U(0, 2)\), \(X_3 = \text{sign}(\theta)(\cos(\theta) - 1)\). This configuration creates a horizontally curled shape in \((X_1, X_3)\), with additional band thickness in the \(X_2\) direction. For \(p > 3\), additional noise dimensions are appended introducing structured, wavy noise.

@@

A latent angle variable \(\theta\) is uniformly sampled from the interval \([0, 2\pi]\). Coordinates in the first two dimensions represent a perfect circle on the plane: \(X_1 = \cos(\theta), \quad X_2 = \sin(\theta)\). For dimensions \(X_3\) through \(X_p\), sinusoidal transformations of the angle \(\theta\) are introduced. The first component is a scaling factor that decreases with the dimension index, defined as \(\text{scale}_j = \sqrt{(0.5)^{j-2}}\) for \(j = 3, \dots, p\). The second component is a phase shift that is proportional to the dimension index, specifically designed to decorrelate the curves, given by the formula \(\phi_j = (j - 2)\pi/2p\). Each additional dimension is computed as: \(X_j = \text{scale}_{j}\sin(\theta + \phi_j), \quad j = 3, \dots, p\).

@@

A latent angle variable \(\theta\) is uniformly sampled from the interval \([0, 2\pi]\). The first three dimensions define a non-circular closed curve, referred to as a ``curvy cycle.'' In this configuration, \(X_1 = \cos(\theta)\) represents horizontal oscillation, while \(X_2 = \sqrt{3}/3 + \sin(\theta)\) introduces a vertical offset to avoid centering the curve at the origin. Additionally, \(X_3 = 1/3\cos(3\theta)\) introduces a third harmonic perturbation that intricately folds the curve three times along its path, creating a unique and complex shape that oscillates in both dimensions while incorporating the effects of the harmonic perturbation.

Together, these define a periodic, non-trivial, closed curve in \(3\text{-}D\) with internal folds that produce a more complex geometry than a standard circle or ellipse. For dimensions \(X_4\) through \(X_p\), additional structured variability is introduced through decreasing amplitude scaling and phase-shifted sine waves. The scaling factor is defined as \(\text{scale}_j = \sqrt{(0.5)^{j-3}}\) for \(j\) ranging from \(4\) to \(p\), which means that the amplitude decreases as the dimension increases. Each dimension \(X_j\) is then calculated using the formula \(X_j = \text{scale}_j\sin(\theta + \phi_j)\), where the phase shift \(\phi_j\) is given by \(\phi_j = (j - 2)\pi/2p\).

@@

Each observation lies on the surface of a sphere in \(3\text{-}D\), constructed by generating \(u \sim U(-1, 1)\) which represents the cosine of the polar angle \(\phi\) and \(\theta \sim U(0, 2\pi)\) which represents the azimuthal angle. The corresponding Cartesian coordinates are calculated as; \(X_1 = r\sqrt{1 - u^2}\cos(\theta)\), \(X_2 = r\sqrt{1 - u^2}\sin(\theta)\), \(X_3 = ru\), which gives points uniformly distributed on the surface of a \(3\text{-}D\) sphere (not within). For \(p > 3\), additional dimensions \(X_4\) to \(X_p\) are generated as low-variance Gaussian noise: \(X_j \sim N(0, 0.05^2),\text{ for }j = 4, \dots, p\).

@@

Each point on the sphere is computed using the spherical-to-Cartesian transformation: \(X_1 = \sin(\phi)\cos(\theta)\), \(X_2 = \sin(\phi)\sin(\theta)\), \(X_3 = \cos(\phi)\). For \(p > 3\), the manifold is embedded in a higher-dimensional space by adding Gaussian noise dimensions: \(X_j \sim N(0, 0.05^2),\text{ for }j = 4, \dots, p\).

@@

Each observation is situated on a restricted \(4\text{-}D\) spherical surface, defined by spherical coordinates. The azimuthal angle \(\theta_1 \sim U(0, \pi)\) in the \((x_1, x_2)\) plane, while the elevation angle \(\theta_2 \sim U(0, \pi)\) in the \((x_2, x_3)\) plane. Additionally, \(\theta_3 \sim U(0, \pi/2)\) in the \((x_3, x_4)\) plane, ensuring that the points remain restricted to a hemisphere. The coordinates are transformed into \(4\text{-}D\) Cartesian space: \(X_1 = \sin(\theta_1)\cos(\theta_2)\), \(X_2 = \sin(\theta_1)\sin(\theta_2)\), \(X_3 = \cos(\theta_1)\cos(\theta_3)\), \(X_4 = \cos(\theta_1)\sin(\theta_3)\). This produces points on one side of a \(4\text{-}D\) unit sphere, effectively generating a \(4\text{-}D\) hemisphere. For \(p > 4\), additional dimensions \(X_5\) to \(X_p\) are added as low-variance Gaussian noise: \(X_j \sim N(0, 0.05^2),\text{ for }j = 5, \dots, p\).

@@

The first two dimensions define a crescent-shaped curve using a semi-circular arc. Let \(\theta \in [\pi/6, 2\pi]\) be a sequence of \(n\) evenly spaced angles. The corresponding \(2\text{-}D\) coordinates are defined by: \(X_1 = \cos(\theta)\), \(X_2 = \sin(\theta)\). For \(p > 2\), additional dimensions \(X_3\) to \(X_p\) are generated using \texttt{gen\_wavydims1()}: \(X_j = s_j\theta + \varepsilon_j,\text{ where }\varepsilon_j \sim N(0, 0.5^2),\text{ and }s_j\text{ is a random scale}\). These noise dimensions are nonlinearly correlated with \(\theta\), producing structured, wave-like variation that aligns with the progression along the crescent curve.

@@

Cartesian coordinates from spherical conversion: \(X_1 = \sin(\phi)\cos(\theta)\), \(X_2 = \sin(\phi)\sin(\theta)\), \(X_3 = \cos(\phi) + \varepsilon\), where \(\varepsilon \sim U(-0.5, 0.5)\) introduces vertical jitter, and \(X_4 = \theta / \max(\theta)\): a normalized progression along the spiral path. This generates a spherical spiral curve embedded in \(4\text{-}D\) space, combining both circular and vertical movement, with gentle curvature and non-linear progression.

If \(p > 4\), the function appends structured, non-linear noise via \texttt{gen\_wavydims2()}. These dimensions are functions of the first coordinate \(X_1\), introducing dependencies that preserve some geometric coherence. Each added dimension follows the form: \(X_j = s_j(-1)^{\lfloor j/2 \rfloor}X_1^{a_j} + \eta_j\), where \(a_j \in \{2, 3, 4, 5\}\) is a randomly chosen polynomial power, \(s_j \sim U(0.5, 2)\) is a scale factor, \(\eta_j \sim U(-\sigma_j, 2\sigma_j)\), with \(\sigma_j \sim U(0, 0.05)\), adds mild randomness.

@@

The first four coordinates follow a \(3\text{-}D\) helix with an additional oscillating component. Let \(\theta \in [0, 5\pi/4]\) be a sequence of angles controlling rotation around a circle. Cartesian coordinates; \(X_1 = \cos(\theta)\): circular trajectory along the x-axis, \(X_2 = \sin(\theta)\): circular trajectory along the y-axis, \(X_3 = 0.05\theta + \varepsilon_3\), with \(\varepsilon_3 \sim U(-0.5, 0.5)\): linear progression (height) with vertical jitter, simulating a helix, and \(X_4 = 0.1\sin(\theta)\): oscillates with \(\theta\), representing a periodic ``wobble'' along the fourth dimension. This results in a helical spiral winding upward along the third axis, with gentle sinusoidal fluctuation in the fourth dimension. When \(p > 4\), the remaining \(p - 4\) dimensions are populated using \texttt{gen\_noisedims()}. These are independent Gaussian noise dimensions: \(X_j \sim N(0, 0.05^2), \quad j = 5, \dots, p\).

@@

The shape is defined by parameter \(\theta \in [0, 2\pi\text{spins}]\), controlling the angular progression of the spiral. The Archimedean spiral in the horizontal plane is represented by; \(X_1 = \theta\cos(\theta)\) for radial expansion in x, and \(X_2 = \theta\sin(\theta)\) for radial expansion in y. The growth pattern resembles a cone, with the height increasing according to \(X_3 = 2\theta / \max(\theta) + \varepsilon_3\), with \(\varepsilon_3 \sim U(-0.1, 0.6).\) Spiral modulation in the fourth dimension is represented by \(X_4 = \theta\sin(2\theta) + \varepsilon_4\), with \(\varepsilon_4 \sim U(-0.1, 0.6)\) which simulates a twisting helical component in a non-radial dimension. This results in a \(3\text{-}D\) spiral surface expanding upward and outward like a cone, with an oscillatory fourth dimension capturing spiral irregularities. For \(p > 4\), the dataset includes isotropic Gaussian noise in the remaining dimensions via \texttt{gen\_noisedims()}: \(X_j \sim N(0, 0.05^2), \quad j = 5, \dots, p\).

@@

The \(X_{1} \sim U(0.1, 2)\): base variable (avoids zero to prevent division errors), \(X_{3} \sim U(0.1, 0.8)\): independent auxiliary variable, \(X_{2} = hc/X_{1} + \text{nonfac}\sin(X_{1})\): non-linear combination of hyperbolic and sinusoidal transformations, creating sharp curvature and oscillation, and \(X_{4} = \cos(\pi X_{1}) + \varepsilon\), with \(\varepsilon \sim U(-0.1, 0.1)\): additional nonlinear variation based on cosine, simulating more subtle periodic structure. These transformations together result in a non-linear surface warped in multiple ways: sharp vertical shifts due to inverse terms, smooth waves from sine and cosine, and additional jitter. If \(p > 4\), the remaining dimensions are populated with Gaussian noise using \texttt{gen\_noisedims()}: \(X_j \sim N(0, 0.05^2), \quad j = 5, \dots, p\).
